---
title: "Class 5: Data Import"
output: github_document
---

```{r setup}
library(tidyverse)
library(readxl)
library(haven)
```

## Read in some data 

Read in the litters dataset. 

```{r}
litters_df = read_csv("./data/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df)
```

## Take a look at the data 

Printing in the console. 

```{r}
litters_df
```

```{r}
head(litters_df)
tail(litters_df)
tail(litters_df, 5)
```

```{r}
skimr::skim(litters_df)
```

Setting NA to missing so weight variables are treated as numeric.

```{r}
litters_df = 
  read_csv(
    file = "./data/FAS_litters.csv",
     na = c(".", "NA", ""))

head(litters_df)
```

```{r}
skimr::skim(litters_df)
```

```{r}
litters_df
names(litters_df)
```


Parsing columns
Giving explicit column specifications. This is done using cols function, where you can give each column a type:

```{r}
litters_df = 
    read_csv(file = "./data/FAS_litters.csv",
        na = c(".", "NA", ""),
    col_types = cols(
      Group = col_character(),
      `Litter Number` = col_character(),
      `GD0 weight` = col_double(),
      `GD18 weight` = col_double(),
      `GD of Birth` = col_integer(),
      `Pups born alive` = col_integer(),
      `Pups dead @ birth` = col_integer(),
      `Pups survive` = col_integer()
    )
  )
tail(litters_df)
```


## Options to read_csv

Check out `?read_csv()` for more information. 

Skips first 10 rows. 

```{r}
litters_df = read_csv("./data/FAS_litters.csv")
```

## Other file formats

Read in an excel file

```{r}
library(readxl)
mlb11_df = read_excel("data/mlb11.xlsx", n_max = 20)
head(mlb11_df, 5)
```

```{r}
mlb_df = read_excel("./data/mlb11.xlsx", range = "A1:F7")
mlb_df
```

Read in a SAS file. 

```{r}
pulse_df = read_sas("./data/public_pulse_data.sas7bdat")
pulse_df
```

## Comparison with Base R 

What about `read.csv` ... ? 

```{r}
litters_base = read.csv("data/FAS_litters.csv")
litters_readr = read_csv("data/FAS_litters.csv")

litters_base
litters_readr
```

## Exporting data 

Export the mlb sub-table. 

```{r}
mlb_df
write_csv(mlb_df, "./data/mlb_subtable.csv")
```

## Learning Assessments

Learning assessment 1: load the FAS_pups.csv dataset 

```{r}
pups_df_rel = read_csv("./data/FAS_pups.csv", skip = 3)
pups_df_rel

pups_df_abs = read_csv("/Users/nickylewis/Desktop/Columbia/School/Fall 2026/Data Science/Class Examples/Class5 - Data Import/data_wrangling_i/data/FAS_pups.csv", skip = 3)
pups_df_abs
```

Learning assessment 2: Repeat the data import process above for the file FAS_pups.csv. Make sure the column names are reasonable, and take some quick looks at the dataset. What happens if your specifications for column parsing aren’t reasonable (e.g. character instead of double, or vice versa)?

```{r}
pups_df = 
  read_csv(
    file = "./data/FAS_pups.csv", 
    skip = 3,
    na = c(".", "NA"), col_types = "fddddd")

skimr::skim(pups_df)
head(pups_df)

```

Learning assessment 3: Import the FAS_pups.csv dataset using read.csv. Compare the class of this dataset to the one produced by read_csv. Try printing both in the console – what happens? After cleaning up the names, try accessing the Sex variable using S (ie. pups_df$S). What happens? 

```{r}
pups_base = read.csv("./data/FAS_pups.csv", skip = 3)
pups_readr = read_csv("./data/FAS_pups.csv", skip = 3)

View(pups_base)
View(pups_readr)

pups_base
pups_readr

pups_base$S
pups_readr$S

```





