---
title: "Class 6: Data Manipulation"
output: github_document
---

```{r setup}
library(tidyverse)
```

## Load in the FAS Litters Data 

```{r}
litters_df = read_csv("./data/FAS_litters.csv", na = c("", "NA", "."))
litters_df = janitor::clean_names(litters_df)
```

## `select`

Choose some columns and not others

a) Choosing specific variables to keep 

```{r}
select(litters_df, group, litter_number)
```

```{r}
select(litters_df, group, gd0_weight)
```

b) Choosing a range of columns / vars to keep 

```{r}
select(litters_df, group, gd0_weight:gd_of_birth)
```

c) Specify which variables you want to lose 

```{r}
select(litters_df, -litter_number)
```

Renaming columns 

```{r}
select(litters_df, GROUP = group, LITTer_NUmBer = litter_number)
```

```{r}
rename(litters_df, GROUP = group, LITTer_NUmBer = litter_number)
```

Select helpers

a) Keeps vars that starts with gd and removes all others 
```{r}
select(litters_df, starts_with("gd"))
```

b) Moving variable to the front, but keeping all other variables
```{r}
select(litters_df, litter_number, everything())
```

```{r}
relocate(litters_df, litter_number)
```

## `filter`

```{r}
filter(litters_df, gd0_weight < 22)
```

```{r}
filter(litters_df, gd0_weight >= 22)
```

```{r}
filter(litters_df, gd_of_birth == 20)
```

```{r}
filter(litters_df, !(gd_of_birth == 20))
```

```{r}
filter(litters_df, gd_of_birth != 20)
```

```{r}
filter(litters_df, gd0_weight >= 22, gd_of_birth == 20)
```

```{r}
filter(litters_df, group == "Mod8")
```

```{r}
filter(litters_df, group %in% c("Con7", "Mod8"))
```

## `mutate`

```{r}
mutate(litters_df, 
       wt_gain = gd18_weight - gd0_weight,
       group = str_to_lower(group))
```

## `arrange`

```{r}
arrange(litters_df, pups_born_alive)
```

```{r}
arrange(litters_df, pups_born_alive, gd0_weight)
```

## `%>%`

One option (intermediate steps) - don't use this
```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv", na = c("", "NA", "."))
litters_clean_name = janitor::clean_names(litters_data_raw)
litters_data_selected = select(litters_clean_name, -pups_survive)
litters_mutated = mutate(litters_data_selected, wt_gain = gd18_weight - gd0_weight)
litters_without_missing = drop_na(litters_mutated, gd0_weight)
```

USE THE PIPE OPERATOR INSTEAD

```{r}
litters_df = 
  read_csv("./data/FAS_litters.csv", na = c("", "NA", ".")) %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(wt_gain = gd18_weight - gd0_weight) %>% 
  drop_na(gd0_weight)
```


|> vs %>%

```{r}
litters_df |>
  lm(wt_gain ~ pups_born_alive, data = _) |>
  broom::tidy()
```


Learning Assessments 

```{r}
pups_df = 
  read_csv(
    "./data/FAS_pups.csv",
    skip = 3,
    na = c("NA", "."))
pups_df = janitor::clean_names(pups_df)
```

1. In the pups data, select the columns containing litter number, sex, and PD ears

```{r}
select(pups_df, litter_number:pd_ears)
```

2a. In the pups data: Filter to include only pups with sex 1

```{r}
filter(pups_df, sex == 1)
```

2b. In the pups data: Filter to include only pups with PD walk less than 11 and sex 2

```{r}
filter(pups_df, sex == 2, pd_walk < 11)
```

3a. In the pups data: Create a variable that subtracts 7 from PD pivot

```{r}
mutate(pups_df, pivot_minus7 = pd_pivot - 7)
```

3b. In the pups data: Create a variable that is the sum of all the PD variables

```{r}
mutate(pups_df, pd_sum = pd_ears + pd_eyes + pd_pivot + pd_walk)
```

4. Write a chain of commands that:
loads the pups data
cleans the variable names
filters the data to include only pups with sex 1
removes the PD ears variable
creates a variable that indicates whether PD pivot is 7 or more days

```{r}
pups_df = 
  read_csv("./data/FAS_pups.csv", na = c("", "NA", "."), skip = 3) |> 
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(pd_pivot_gt7 = pd_pivot > 7)
pups_df
```

