---
title: "Class 7: Tidy Data"
output: github_document
---

```{r setup}
library(tidyverse)
library(dplyr)
```

## `pivot_longer`

Load the PULSE data

```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names()
```

Wide format to long format 

```{r}
pulse_data_tidy = 
  pulse_data |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  )
pulse_data_tidy
```

Rewrite, combine, and extent (to add a mutate step)

```{r}
pulse_data = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) |> 
  relocate(id, visit) |> 
  mutate(visit = recode(visit, "bl" = "00m"))
```


## `pivot_wider`

Make up some data. 

```{r}
analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"), 
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 3.5, 4)
  )
analysis_result |> 
  pivot_wider(
    names_from = "time", 
    values_from = "mean"
  )
```


## Binding rows

Using the LotR data. 

First step: import each table 

```{r}
fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") |> 
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") |> 
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") |> 
  mutate(movie = "return_king")
```

Bind all the rows together 

```{r}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) |> 
  janitor::clean_names() |> 
  relocate(movie) |> 
  pivot_longer(
    female:male, 
    names_to = "gender",
    values_to = "words"
  )
```


## Joining datasets

Import and clean the FAS datasets. 

```{r}
pups_df = 
  read_csv("./data/FAS_pups.csv", skip = 3, na = c(".", "NA")) |> 
  janitor::clean_names() |> 
  mutate(sex = recode(sex, `1` = "male", `2` = "female"))

litters_df = 
  read_csv("./data/FAS_litters.csv") |> 
  janitor::clean_names() |> 
  relocate(litter_number) |> 
  separate(group, into = c("dose", "day_of_tx"), sep = 3)
```

Join the FAS datasets
Left join; merge litters data into the pups data 

```{r}
fas_df = 
  left_join(pups_df, litters_df, by = "litter_number") |> 
  arrange(litter_number) |> 
  relocate(litter_number, dose, day_of_tx)
```


## Learning assessments

1. pivot_longer
In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?

```{r}
litters_wide = 
  read_csv(
    "./data/FAS_litters.csv",
    na = c(".", "NA", "")) |> 
  janitor::clean_names() |> 
  select(litter_number, ends_with ("weight")) |> 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd", 
    values_to = "weight") |> 
  mutate(
    gd = case_match(
      gd, 
      "gd0_weight" ~ 0,
      "gd18_weight" ~ 18,
    )
  )
  
```






